Mental Health Factors and Covid

### 2025-05-02 code/analysis-code/analysis.qmd
Part 6 updates

Next, we will make the multivariate model. I will use cross validation for this.Pooled AUC is calculated and a observed-expected plot is generated.
```{r}
library(pROC)
library(dplyr)

cross_validation_imputed_auc_logit <- function(imputed_data, predictors, k_folds = 5) {
  auc_list <- numeric(length(imputed_data))  # Store AUC for each imputation

@@

```{r}
library(glmnet)
library(caret)
library(Metrics)  # for rmse()
library(dplyr)

set.seed(1234)
k <- 5
auc_list <- c()

@@

pooled_auc <- mean(auc_list)
print(pooled_auc)

@@

# Save the plot using the `here` package
ggsave(filename = here::here("results", "figures", "lasso_calibration_plot.png"),
       plot = Lasso_plot,
       width = 8, height = 6, dpi = 300)

################################################################################
# Keep variable names for predictors
formula <- as.formula("COVID ~ .")

# Explicitly remove rows with any NA in predictors (even if COVID is not NA)
train_data <- train_data %>% select(COVID, everything()) %>% drop_na()

# Now get model matrix
x_train <- model.matrix(formula, data = train_data)[, -1]
y_train <- train_data$COVID

nrow(x_train)
length(y_train)
```

```{r}
# Load required libraries
library(glmnet)      # For Lasso regression
library(pROC)        # For AUC calculation
library(ggplot2)     # For plotting
library(dplyr)       # For data manipulation
library(mice)        # For working with imputed data
library(tidyr)       # For data reshaping

# --- LASSO MODEL WITH CROSS-VALIDATION ON IMPUTED DATA ---

# Function to run Lasso with sample weights and calculate performance metrics
run_lasso_cv <- function(imp_data, outcome = "COVID", 
                        predictors = c("SLEEP", "DEPRESSED", "SOCIAL", "TROUBLE_SLEEPING", "EXERCISE"),
                        weight_var = "sampleweight") {

  # Extract data from current imputation
  X <- as.matrix(imp_data[, predictors])
  y <- imp_data[[outcome]]
  weights <- imp_data[[weight_var]]

  # Standardize predictors (important for Lasso)
  X_scaled <- scale(X)

  # Run cross-validated Lasso with sample weights
  set.seed(123) # For reproducibility
  cv_model <- cv.glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,       # 1 = Lasso
    weights = weights,
    nfolds = 5,
    type.measure = "auc"
  )

  # Get the best lambda (penalty parameter)
  best_lambda <- cv_model$lambda.min

  # Refit model with best lambda
  final_model <- glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,
    weights = weights,
    lambda = best_lambda
  )

  # Get predictions
  pred_probs <- predict(final_model, newx = X_scaled, type = "response")

  # Calculate AUC
  roc_obj <- roc(y, pred_probs)
  auc_value <- auc(roc_obj)

  # Create calibration data for expected vs observed plot
  n_groups <- 10  # Number of bins for calibration plot

  # Create calibration data
  calib_data <- data.frame(
    predicted = as.vector(pred_probs),
    observed = y
  )

  # Sort by predicted probability and divide into bins
  calib_data <- calib_data[order(calib_data$predicted), ]
  calib_data$bin <- cut(seq(1, nrow(calib_data)), breaks = n_groups, labels = FALSE)

  # Calculate mean predicted and observed values for each bin
  calibration <- calib_data %>%
    group_by(bin) %>%
    summarize(
      n = n(),
      mean_predicted = mean(predicted),
      mean_observed = mean(observed)
    )

  # Return results
  return(list(
    model = final_model,
    lambda = best_lambda,
    auc = auc_value,
    calibration = calibration,
    coefs = coef(final_model)
  ))
}

# ----------------------------
# MAIN ANALYSIS
# ----------------------------

# Check the structure of the imputed data
print(class(data_imputed))

# Initialize results containers
auc_values <- numeric(length = 5)
calibration_results <- list()
coef_results <- list()

# Function to run Lasso with sample weights and calculate performance metrics
run_lasso_cv <- function(imp_data, outcome = "COVID", 
                        predictors = c("SLEEP", "DEPRESSED", "SOCIAL", "TROUBLE_SLEEPING", "EXERCISE"),
                        weight_var = "sampleweight") {

  # Extract data from current imputation and handle missing values
  # First, check for and handle missing data
  complete_cases <- complete.cases(imp_data[, c(outcome, predictors, weight_var)])
  if (!all(complete_cases)) {
    message("Removing ", sum(!complete_cases), " rows with missing values")
    imp_data <- imp_data[complete_cases, ]
  }

  X <- as.matrix(imp_data[, predictors])
  y <- imp_data[[outcome]]
  weights <- imp_data[[weight_var]]

  # Standardize predictors (important for Lasso)
  X_scaled <- scale(X)

  # Run cross-validated Lasso with sample weights
  set.seed(123) # For reproducibility
  cv_model <- cv.glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,       # 1 = Lasso
    weights = weights,
    nfolds = 5,
    type.measure = "auc"
  )

  # Get the best lambda (penalty parameter)
  best_lambda <- cv_model$lambda.min

  # Refit model with best lambda
  final_model <- glmnet(
    x = X_scaled,
    y = y,
    family = "binomial",
    alpha = 1,
    weights = weights,
    lambda = best_lambda
  )

# Combine calibration results from all imputations
all_calibration <- do.call(rbind, calibration_results)

# ----------------------------
# OUTPUT RESULTS
# ----------------------------

# Print AUC values for each imputation
cat("\nAUC values for each imputation:\n")
print(auc_values)
cat("\nMean AUC across imputations:", mean(auc_values), "\n")

# Plot expected vs observed probabilities for all imputations
calibration_plot <- ggplot(all_calibration, aes(x = mean_predicted, y = mean_observed, color = factor(imputation))) +
  geom_point(size = 3) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Expected vs. Observed Probabilities",
    subtitle = paste("Mean AUC =", round(mean(auc_values), 3)),
    x = "Expected (Predicted Probability)",
    y = "Observed (Actual Proportion)",
    color = "Imputation"
  ) +
  theme_minimal() +
  coord_equal() +
  xlim(0, 1) + 
  ylim(0, 1)

print(calibration_plot)

# Print average coefficients across imputations
cat("\nAverage Lasso coefficients across imputations:\n")
# Extract coefficient names from the first imputation
coef_names <- rownames(coef_results[[1]])

# Create matrix to store coefficients from all imputations
all_coefs <- matrix(0, nrow = length(coef_names), ncol = 5)
rownames(all_coefs) <- coef_names

# Fill in the coefficient matrix
for (i in 1:5) {
  all_coefs[, i] <- as.vector(coef_results[[i]])
}

# Calculate mean coefficients
mean_coefs <- rowMeans(all_coefs)
coef_summary <- data.frame(
  Variable = coef_names,
  Mean_Coefficient = mean_coefs
)

# Print non-zero coefficients
non_zero_coefs <- coef_summary[abs(coef_summary$Mean_Coefficient) > 0.001, ]
print(non_zero_coefs[order(abs(non_zero_coefs$Mean_Coefficient), decreasing = TRUE), ])
```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

calib_data_list <- vector("list", length(data_imputed))

for (i in seq_along(data_imputed)) {
  dat <- data_imputed[[i]] %>%
    drop_na(all_of(c(outcome_var, predictor_vars, "SAMPWEIGHT")))

  X <- model.matrix(as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))), dat)[, -1]
  y <- dat[[outcome_var]]
  w <- dat$SAMPWEIGHT

  set.seed(1234 + i)
  train_idx <- sample(seq_len(nrow(X)), size = 0.7 * nrow(X))
  test_idx <- setdiff(seq_len(nrow(X)), train_idx)

  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  w_train <- w[train_idx]

  X_test <- X[test_idx, ]
  y_test <- y[test_idx]

  cv_fit <- cv.glmnet(
    x = X_train,
    y = y_train,
    weights = w_train,
    alpha = 1,
    family = "binomial",
    standardize = TRUE
  )

  pred_probs <- predict(cv_fit, newx = X_test, s = "lambda.min", type = "response") %>% as.vector()

  # Count unique predicted probabilities
  n_unique_preds <- length(unique(pred_probs))

  # Only proceed with binning if there are at least 2 unique predicted probabilities
  if (n_unique_preds >= 2) {
    # Safely bin the predicted probabilities
    n_bins <- min(5, n_unique_preds)

    calib_data <- tibble(
      actual = y_test,
      predicted = pred_probs
    ) %>%
      mutate(bin = cut_number(predicted, n = n_bins, labels = FALSE)) %>%
      group_by(bin) %>%
      summarise(
        mean_pred = mean(predicted),
        obs_rate = mean(actual),
        .groups = "drop"
      ) %>%
      mutate(imputation = paste0("Imputation ", i))

    calib_data_list[[i]] <- calib_data
  } else {
    message("Skipping calibration plot for imputation ", i, ": too few unique predicted values (", n_unique_preds, " unique values)")
  }
}

# Combine all calibration data
all_calib_data <- bind_rows(calib_data_list)

# Create calibration plot
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()

# Combine calibration data
all_calib_data <- bind_rows(calib_data_list)

# Plot calibration curve
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()

# Combine calibration data
all_calib_data <- bind_rows(calib_data_list)

# Plot calibration curve
ggplot(all_calib_data, aes(x = mean_pred, y = obs_rate, color = imputation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous("Expected Probability", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous("Observed Proportion", limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  labs(title = "Calibration Plot: Lasso Logistic Regression", color = "Imputation") +
  theme_minimal()
# Save the plot
ggsave(filename = here::here("results", "figures", "lasso_calibration_plot.png"),
       plot = calibration_plot_lasso,
       width = 8, height = 6, dpi = 300)

# Initialize list to store AUCs
auc_list <- vector("numeric", length(data_imputed))

for (i in seq_along(data_imputed)) {
  dat <- data_imputed[[i]] %>%
    drop_na(all_of(c(outcome_var, predictor_vars, "SAMPWEIGHT")))

  X <- model.matrix(as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = "+"))), dat)[, -1]
  y <- dat[[outcome_var]]
  w <- dat$SAMPWEIGHT

  set.seed(1234 + i)
  train_idx <- sample(seq_len(nrow(X)), size = 0.7 * nrow(X))
  test_idx <- setdiff(seq_len(X), train_idx)

  X_train <- X[train_idx, ]
  y_train <- y[train_idx]
  w_train <- w[train_idx]

  X_test <- X[test_idx, ]
  y_test <- y[test_idx]

  # Fit lasso model
  cv_fit <- cv.glmnet(
    x = X_train,
    y = y_train,
    weights = w_train,
    alpha = 1,
    family = "binomial",
    standardize = TRUE
  )

  # Predict probabilities
  pred_probs <- predict(cv_fit, newx = X_test, s = "lambda.min", type = "response") %>% as.vector()

  # Compute AUC
  auc_val <- roc(y_test, pred_probs)$auc
  auc_list[i] <- auc_val
}

# Print AUCs
cat("AUC for each imputation:\n")
print(round(auc_list, 3))

# Mean (pooled) AUC
cat("\nPooled AUC across imputations:", round(mean(auc_list), 3), "\n")
```

@@

  list(
    auc_per_imputation = auc_list,
    overall_auc = mean(auc_list),
    predictions = predictions_list  # List of tibbles with actual + predicted
  )
}
# Assuming you have your imputed_data and a vector of predictors
results_with_preds <- cross_validation_imputed_predictions(
  imputed_data = data_imputed,
  predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL")
)

# View results
results$auc_per_imputation  # AUC for each imputed dataset
results$overall_auc         # Average AUC across imputations

@@

Now, we will make expected versus observed plots.
```{r}
plot_calibration_multiple <- function(actual, predicted_list, n_bins = 10, title = "Calibration Plot") {
  # Combine calibration data across imputations
  calib_data_all <- map2_dfr(predicted_list, seq_along(predicted_list), function(predicted, i) {
    tibble(actual = actual, predicted = predicted) %>%
      mutate(bin = ntile(predicted, n_bins)) %>%
      group_by(bin) %>%
      summarise(
        mean_pred = mean(predicted),
        obs_rate = mean(actual),
        .groups = "drop"
      ) %>%
      mutate(imputation = paste0("Imputation ", i))
  })

  # Plot
  ggplot(calib_data_all, aes(x = mean_pred, y = obs_rate, color = imputation)) +
    geom_line() +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    scale_x_continuous("Expected probability", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    scale_y_continuous("Observed proportion", limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(title = title, color = "Imputation") +
    theme_minimal()
}

# Example usage:
plot_cal <- plot_calibration_multiple(
  actual = test_data$COVID,
  predicted_list = pred_probs_list,  # this should be a list of 5 vectors
  title = "Random Forest Calibration Across 5 Imputations"
)

# Display plot
print(plot_cal)

# Save it
ggsave(
  filename = here("results", "figures", "rf_calibration_all_imputations.png"),
  plot = plot_cal,
  width = 8,
  height = 6,
  dpi = 300
)

>>>>

Next, we will make the multivariate model. I will use cross validation for this. Pooled AUC is calculated and a observed-expected plot is generated.
```{r}
cross_validation_imputed_auc_logit <- function(imputed_data, predictors, k_folds = 5) {
  auc_list <- numeric(length(imputed_data))  # Store AUC for each imputation

@@

```{r}
set.seed(1234)
k <- 5
auc_list <- c()

@@

auc_list
pooled_auc <- mean(auc_list)
print(pooled_auc)

@@

# Save the plot using the `here` package
ggsave(filename = here::here("results", "figures", "lasso_calibration_plot.png"),
       plot = Lasso_plot,
       width = 8, height = 6, dpi = 300)
```

@@

  list(
    auc_per_imputation = auc_list,
    overall_auc = mean(auc_list),
    predictions = predictions_list  # List of tibbles with actual + predicted
  )
}
# Assuming you have your imputed_data and a vector of predictors
results_with_preds2 <- cross_validation_imputed_predictions(
  imputed_data = data_imputed,
  predictors = c("EXERCISE", "SLEEP", "DEPRESSED", "TROUBLE_SLEEPING", "SOCIAL")
)

# View results
results_with_preds2$auc_per_imputation  # AUC for each imputed dataset
results_with_preds2$overall_auc         # Average AUC across imputations

@@

Now, we will make a plot of the AUCs.
```{r}
# Step 1: Combine AUCs into a tidy data frame
auc_df <- data.frame(
  model = rep(c("Logistic Regression", "Lasso", "Random Forest"), each = 5),
  imputation = rep(1:5, times = 3),
  AUC = c(
    results_logit$auc_per_imputation,
    auc_list,
    results_with_preds2$auc_per_imputation
  )
)

# Step 2: Calculate mean AUC per model
auc_means <- auc_df %>%
  group_by(model) %>%
  summarise(mean_auc = mean(AUC), .groups = "drop")

# Step 3: Plot
ggplot(auc_df, aes(x = model, y = AUC, color = model)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +  # dots for each imputation
  geom_point(data = auc_means, aes(x = model, y = mean_auc),
             color = "black", size = 4) +             # black dot for mean
  geom_text(data = auc_means, aes(x = model, y = mean_auc, 
             label = formatC(mean_auc, format = "f", digits = 4)),
            vjust = -1.2, color = "black", size = 4.5) +  # label for mean
  labs(
    title = "AUCs by Model Type Across 5 Imputations",
    x = "Model",
    y = "AUC"
  ) +
  theme_minimal(base_size = 14) +
  scale_y_continuous(limits = c(0.45, 0.6), breaks = seq(0.45, 0.6, 0.05)) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")

#save plot
ggsave(
  filename = here("results", "figures", "auc_comparison_plot.png"),  # save in outputs/ directory
  plot = last_plot(),       # or specify your plot object explicitly
  width = 8, height = 6,    # size in inches
  dpi = 300                 # high-quality resolution
)
```

### 2025-05-02 products/manuscript/Manuscript.qmd
Part 6 updates

To evalaute the performance of these models, the area under the curve (AUC) was calculated. This metric was chosen due to it's discrimination between binary outcome of the COVID-19 variable. Higher AUC indicates better model preformance.

```{r}
#| label: fig-aucplot
#| fig-cap: "AUC comparison plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "auc_plot.png"))
```

>>>>

To evalaute the performance of these models, the area under the curve (AUC) was calculated. This metric was chosen due to it's discrimination between binary outcome of the COVID-19 variable. Higher AUC indicates better model preformance. Desspite the Logistic Regression model and the Lasso Regression model have the same mean AUC, they are not the same, shown by the differing values for the imputations. Code was checked to ensure the models were not the same.

```{r}
#| label: fig-aucplot
#| fig-cap: "AUC comparison plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "auc_comparison_plot.png"))
```

Observed vs. expected plots for the three models compare the predictive power of the models from the cross validated data. Training data was used for expectations and observed data was drawn from the testing data. Plots evaluate the models' modest predictive power on COVID-19 test outcomes from mental health variables.

```{r}
#| label: fig-log-plot
#| fig-cap: "Logistic Regression calibration plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "log_calibration_plot.png"))
```

```{r}
#| label: fig-lasso-plot
#| fig-cap: "LASSO Regression calibration plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "lasso_calibration_plot.png"))
```

```{r}
#| label: fig-rf-plot
#| fig-cap: "Random Forest calibration plot"
#| echo: FALSE
here()
knitr::include_graphics(here("results","figures", "rf_calibration_all_imputations.png"))
```
